---
title: "Regression Diagnostics and Model Evaluation of Housing Prices"
author: "Isuri Nawodya Herath S W M M G"
date: "2026-01-13"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    highlight: tango
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center",
  fig.width = 10,
  fig.height = 6
)
options(scipen = 999)
```

---

# Introduction

Regression analysis serves as a foundational tool in predictive analytics, enabling researchers and practitioners to understand relationships between variables and make informed predictions. This report presents a comprehensive regression analysis of the Ames Housing dataset, which contains detailed information about residential property sales in Ames, Iowa, from 2006 to 2010.

The primary objective is to develop, evaluate, and refine multiple linear regression models capable of accurately predicting housing sale prices based on various property characteristics. The analysis emphasizes identification and correction of common regression issues including non-linearity, multicollinearity, outliers, and influential observations.

---

# Setup

## Load Packages

```{r load-packages}
# install.packages(c("tidyverse", "car", "corrplot", "ggcorrplot", "leaps"))

library(tidyverse)
library(car)
library(corrplot)
library(ggcorrplot)
library(leaps)
```

## Load Data

```{r load-data}
ames_data <- read.csv("AmesHousing.csv")

# Clean column names by removing dots
names(ames_data) <- gsub("\\.", "", names(ames_data))

# Remove Order column
ames_data <- ames_data %>% select(-Order)

glimpse(ames_data)
```

---

# Exploratory Data Analysis

## Variable Types

```{r variable-types}
numeric_vars <- ames_data %>% select_if(is.numeric)
categorical_vars <- ames_data %>% select_if(is.character)

cat("Numeric variables:", ncol(numeric_vars), "\n")
cat("Categorical variables:", ncol(categorical_vars), "\n")
cat("Total observations:", nrow(ames_data), "\n")
```

## Missing Values

```{r missing-values}
missing_summary <- data.frame(
  Variable = names(ames_data),
  Missing_Count = colSums(is.na(ames_data)),
  Missing_Percent = round(colSums(is.na(ames_data)) / nrow(ames_data) * 100, 2)
)
missing_summary <- missing_summary[missing_summary$Missing_Count > 0, ]
missing_summary <- missing_summary[order(-missing_summary$Missing_Count), ]
print(missing_summary)
```

## Distributions of Key Numeric Variables

```{r distributions, fig.height=8}
key_vars <- c("SalePrice", "GrLivArea", "LotArea", "TotalBsmtSF",
              "GarageArea", "YearBuilt", "OverallQual", "OverallCond")
key_vars <- key_vars[key_vars %in% names(ames_data)]

par(mfrow = c(3, 3), mar = c(3, 3, 2, 1))
for (var in key_vars) {
  hist(ames_data[[var]],
       main = paste("Distribution of", var),
       xlab = var,
       col = "steelblue",
       breaks = 30,
       cex.main = 0.9)
}
par(mfrow = c(1, 1))
```

The histograms reveal several important distributional characteristics:

- **SalePrice** - right-skewed with most properties selling between \$100,000 and \$250,000
- **GrLivArea** - approximately normal, centered around 1,500 sq ft
- **LotArea** - severely right-skewed with extreme outliers exceeding 150,000 sq ft
- **YearBuilt** - bimodal pattern with peaks around 1960 and 2000–2010
- **OverallQual** - approximately normal, centered at ratings 5–6

## Outlier Detection via Boxplots

```{r boxplots, fig.height=5}
par(mfrow = c(1, 4), mar = c(3, 3, 2, 1))
boxplot_vars <- c("SalePrice", "GrLivArea", "LotArea", "TotalBsmtSF")
for (var in boxplot_vars) {
  boxplot(ames_data[[var]],
          main = var,
          col = "lightgreen",
          cex.main = 0.8)
}
par(mfrow = c(1, 1))
```

## SalePrice Detailed Analysis

```{r saleprice-analysis, fig.height=5}
# Summary statistics
saleprice_stats <- ames_data %>%
  summarise(
    Mean   = round(mean(SalePrice, na.rm = TRUE), 2),
    Median = round(median(SalePrice, na.rm = TRUE), 2),
    SD     = round(sd(SalePrice, na.rm = TRUE), 2),
    Min    = round(min(SalePrice, na.rm = TRUE), 2),
    Max    = round(max(SalePrice, na.rm = TRUE), 2),
    Q1     = round(quantile(SalePrice, 0.25, na.rm = TRUE), 2),
    Q3     = round(quantile(SalePrice, 0.75, na.rm = TRUE), 2)
  )
print(saleprice_stats)

# Histogram and boxplot side by side
par(mfrow = c(1, 2))
hist(ames_data$SalePrice,
     main = "Distribution of Sale Prices",
     xlab = "Sale Price ($)",
     col = "lightblue",
     breaks = 50)
boxplot(ames_data$SalePrice,
        main = "Boxplot of Sale Prices",
        ylab = "Sale Price ($)",
        col = "lightgreen")
par(mfrow = c(1, 1))
```

```{r saleprice-qqplot, fig.height=5}
# Q-Q plot and density plot
par(mfrow = c(1, 2))
qqnorm(ames_data$SalePrice, main = "Q-Q Plot: SalePrice")
qqline(ames_data$SalePrice, col = "red")

plot(density(ames_data$SalePrice, na.rm = TRUE),
     main = "Density Plot: SalePrice",
     xlab = "Sale Price ($)",
     col = "darkblue",
     lwd = 2)
par(mfrow = c(1, 1))
```

The median sale price (~\$160,000) lies below the mean (~\$180,921), confirming positive skewness. The Q-Q plot shows the upper tail curves sharply upward, indicating heavy right-tail behaviour consistent with luxury property outliers.

---

# Correlation Analysis

## Data Preparation - Impute Missing Values

```{r impute}
# Create HouseAge variable
ames_clean <- ames_data

if ("YearBuilt" %in% names(ames_clean) && "YrSold" %in% names(ames_clean)) {
  ames_clean$HouseAge <- ames_clean$YrSold - ames_clean$YearBuilt
}

# Remove variables with >30% missing
ames_clean <- ames_clean %>%
  select(-Alley, -FireplaceQu, -PoolQC, -Fence, -MiscFeature)

# Impute numeric missing values with column mean
numeric_columns <- names(ames_clean)[sapply(ames_clean, is.numeric)]
for (col in numeric_columns) {
  if (sum(is.na(ames_clean[[col]])) > 0) {
    ames_clean[[col]][is.na(ames_clean[[col]])] <- mean(ames_clean[[col]], na.rm = TRUE)
  }
}

cat("Missing values remaining:", sum(colSums(is.na(ames_clean))), "\n")
```

## Correlation Matrix

```{r correlation-matrix, fig.height=8}
numeric_vars <- ames_clean %>% select_if(is.numeric) %>% select(-PID)
cor_matrix <- cor(numeric_vars, use = "complete.obs")

# Top correlations with SalePrice
saleprice_cor <- cor_matrix[, "SalePrice"]
saleprice_cor_sorted <- sort(abs(saleprice_cor), decreasing = TRUE)
print(head(saleprice_cor_sorted, 15))
```

```{r corrplot, fig.height=7}
# Focused correlation plot for key variables
key_numeric_vars <- numeric_vars %>%
  select(SalePrice, GrLivArea, TotalBsmtSF, GarageArea,
         YearBuilt, OverallQual, OverallCond, LotArea, OpenPorchSF)

cor_matrix_key <- cor(key_numeric_vars, use = "complete.obs")

corrplot(
  cor_matrix_key,
  method = "circle",
  type = "upper",
  tl.col = "black",
  tl.srt = 45,
  tl.cex = 0.8,
  title = "Correlation Matrix of Selected Key Numeric Variables",
  mar = c(0, 0, 1, 0)
)
```

Key correlations with `SalePrice`: `OverallQual` (r = 0.80), `GrLivArea` (r = 0.70), `GarageArea` (r = 0.64), `TotalBsmtSF` (r = 0.63), `YearBuilt` (r = 0.56).

## Scatterplot Comparison

```{r scatterplots}
# Identify highest, lowest, and moderate correlations
saleprice_cor_df <- data.frame(
  Variable = names(saleprice_cor),
  Correlation = saleprice_cor
) %>%
  filter(Variable != "SalePrice") %>%
  arrange(desc(abs(Correlation)))

highest_cor_var <- saleprice_cor_df$Variable[1]
lowest_cor_var  <- saleprice_cor_df$Variable[nrow(saleprice_cor_df)]
mid_cor_var     <- saleprice_cor_df$Variable[which.min(abs(abs(saleprice_cor_df$Correlation) - 0.5))]

# Highest correlation
ggplot(numeric_vars, aes(x = .data[[highest_cor_var]], y = SalePrice)) +
  geom_point(color = "blue", alpha = 0.5) +
  geom_smooth(method = "lm", se = FALSE, color = "#A32424") +
  labs(title = paste("SalePrice vs", highest_cor_var,
                     "(Highest Correlation: r =", round(saleprice_cor_df$Correlation[1], 2), ")"),
       x = highest_cor_var, y = "Sale Price ($)") +
  theme_minimal()
```

```{r scatterplot-lowest}
# Lowest correlation
ggplot(numeric_vars, aes(x = .data[[lowest_cor_var]], y = SalePrice)) +
  geom_point(color = "#00832D", alpha = 0.5) +
  geom_smooth(method = "lm", se = FALSE, color = "#A32424") +
  labs(title = paste("SalePrice vs", lowest_cor_var, "(Lowest Correlation)"),
       x = lowest_cor_var, y = "Sale Price ($)") +
  theme_minimal()
```

```{r scatterplot-mid}
# Moderate correlation
ggplot(numeric_vars, aes(x = .data[[mid_cor_var]], y = SalePrice)) +
  geom_point(color = "purple", alpha = 0.5) +
  geom_smooth(method = "lm", se = FALSE, color = "#A32424") +
  labs(title = paste("SalePrice vs", mid_cor_var, "(Moderate Correlation: r ≈ 0.50)"),
       x = mid_cor_var, y = "Sale Price ($)") +
  theme_minimal()
```

---

# Initial Regression Model (Model 1)

## Fit the Model

```{r model1}
model1 <- lm(SalePrice ~ OverallQual + GrLivArea + GarageArea + HouseAge,
             data = ames_clean)

summary(model1)
```

## Model Equation

```{r model1-equation}
coef_model1 <- coef(model1)

cat("SalePrice =", round(coef_model1[1], 2),
    "+", round(coef_model1[2], 2), "* OverallQual",
    "+", round(coef_model1[3], 2), "* GrLivArea",
    "+", round(coef_model1[4], 2), "* GarageArea",
    "+", round(coef_model1[5], 2), "* HouseAge\n")
```

## Coefficient Interpretations

- **OverallQual** - each 1-unit increase in quality rating is associated with ~\$23,744 increase in sale price
- **GrLivArea** - each additional square foot of living area increases sale price by ~\$55.81
- **GarageArea** - each additional square foot of garage space increases sale price by ~\$60.77
- **HouseAge** - each additional year of age decreases sale price by ~\$380.84

---

# Diagnostic Evaluation

## Diagnostic Plots

```{r diagnostics-model1, fig.height=8}
par(mfrow = c(2, 2))
plot(model1, which = 1, pch = 21, bg = "#33ccff", col = "#004466", cex = 0.8)
plot(model1, which = 2, pch = 21, bg = "#33ccff", col = "#004466", cex = 0.8)
plot(model1, which = 3, pch = 21, bg = "#33ccff", col = "#004466", cex = 0.8)
plot(model1, which = 5, pch = 21, bg = "#33ccff", col = "#004466", cex = 0.8)
par(mfrow = c(1, 1))
```

- **Residuals vs Fitted** - mild funnel shape at higher fitted values suggesting heteroscedasticity
- **Q-Q Plot** - approximate normality in center; heavy right tail deviation
- **Scale-Location** - slight upward trend confirming mild heteroscedasticity
- **Residuals vs Leverage** - observations 1499, 2181, 2182 near/beyond Cook's distance 0.5 contour

## Multicollinearity Assessment (VIF)

```{r vif}
vif_values <- vif(model1)
print(vif_values)

if (any(vif_values > 10)) {
  cat("SEVERE multicollinearity detected - VIF > 10\n")
} else if (any(vif_values > 5)) {
  cat("MODERATE multicollinearity detected - VIF > 5\n")
} else {
  cat("No significant multicollinearity - all VIF values < 5\n")
}
```

## Outlier Analysis

```{r outliers}
# Standardized residuals
std_residuals <- rstandard(model1)
outliers_index <- which(abs(std_residuals) > 3)
cat("Outliers with |standardized residual| > 3:", length(outliers_index), "\n")

# Cook's Distance
cooks_d <- cooks.distance(model1)
influential <- which(cooks_d > 4 / nrow(ames_clean))
cat("Influential points (Cook's D > 4/n):", length(influential), "\n")
```

```{r cooks-plot}
plot(cooks_d, type = "h", col = "steelblue",
     main = "Cook's Distance",
     xlab = "Observation Index",
     ylab = "Cook's Distance")
abline(h = 4 / nrow(ames_clean), col = "red", lty = 2)
```

Three observations stand out dramatically - index ~1500 (Cook's D > 0.6), ~1700, and ~2200. These properties had above-grade living areas exceeding 4,000 sq ft but sold for prices much lower than predicted, likely representing partial sales or unusual circumstances.

---

# Improved Model (Model 2)

## Remove Outliers and Refit

```{r model2}
# Remove properties with GrLivArea > 4000 per dataset documentation
ames_clean2 <- ames_clean[ames_clean$GrLivArea <= 4000, ]
cat("Removed", nrow(ames_clean) - nrow(ames_clean2), "observations with GrLivArea > 4000\n")

# Refit the model
model2 <- lm(SalePrice ~ OverallQual + GrLivArea + GarageArea + HouseAge,
             data = ames_clean2)

summary(model2)
```

## Diagnostic Plots - Model 2

```{r diagnostics-model2, fig.height=8}
par(mfrow = c(2, 2))
plot(model2, which = 1, pch = 21, bg = "lightgreen", col = "darkgreen", cex = 0.8)
plot(model2, which = 2, pch = 21, bg = "lightgreen", col = "darkgreen", cex = 0.8)
plot(model2, which = 3, pch = 21, bg = "lightgreen", col = "darkgreen", cex = 0.8)
plot(model2, which = 5, pch = 21, bg = "lightgreen", col = "darkgreen", cex = 0.8)
par(mfrow = c(1, 1))
```

## Model 1 vs Model 2 Comparison

```{r model-comparison}
model_comparison <- data.frame(
  Metric = c("R-squared", "Adjusted R-squared", "Residual Std Error", "Observations"),
  Model_1 = c(
    round(summary(model1)$r.squared, 7),
    round(summary(model1)$adj.r.squared, 7),
    round(summary(model1)$sigma, 7),
    nobs(model1)
  ),
  Model_2 = c(
    round(summary(model2)$r.squared, 7),
    round(summary(model2)$adj.r.squared, 7),
    round(summary(model2)$sigma, 7),
    nobs(model2)
  )
)
print(model_comparison)

# Improvement check
if (summary(model2)$adj.r.squared > summary(model1)$adj.r.squared) {
  cat("Model 2 shows an improvement with a higher Adjusted R-squared than Model 1.\n")
} else {
  cat("Model 2 does not improve upon Model 1.\n")
}
```

---

# All Subsets Regression

## Run Subsets Regression

```{r subsets}
predictors <- c("OverallQual", "GrLivArea", "GarageArea", "HouseAge",
                "TotalBsmtSF", "X1stFlrSF", "FullBath", "TotRmsAbvGrd",
                "Fireplaces", "LotArea")
predictors <- predictors[predictors %in% names(ames_clean2)]

formula_str <- paste("SalePrice ~", paste(predictors, collapse = " + "))

regfit_full <- regsubsets(as.formula(formula_str),
                          data = ames_clean2,
                          nvmax = length(predictors))

reg_summary <- summary(regfit_full)

best_cp   <- which.min(reg_summary$cp)
best_bic  <- which.min(reg_summary$bic)
best_adjr2 <- which.max(reg_summary$adjr2)

cat("Best model by Cp:", best_cp, "variables\n")
cat("Best model by BIC:", best_bic, "variables\n")
cat("Best model by Adjusted R²:", best_adjr2, "variables\n")
```

## Selection Criteria Plots

```{r subsets-plots, fig.height=8}
par(mfrow = c(2, 2))
plot(reg_summary$cp,    xlab = "Number of Variables", ylab = "Cp",           type = "b")
points(best_cp,    reg_summary$cp[best_cp],       col = "red", cex = 2, pch = 20)

plot(reg_summary$bic,   xlab = "Number of Variables", ylab = "BIC",          type = "b")
points(best_bic,   reg_summary$bic[best_bic],     col = "red", cex = 2, pch = 20)

plot(reg_summary$adjr2, xlab = "Number of Variables", ylab = "Adjusted R²",  type = "b")
points(best_adjr2, reg_summary$adjr2[best_adjr2], col = "red", cex = 2, pch = 20)
par(mfrow = c(1, 1))
```

## Fit Preferred Model

```{r preferred-model}
best_model_vars <- names(coef(regfit_full, best_bic))[-1]
cat("Best model variables (by BIC):\n")
print(best_model_vars)

formula_best <- paste("SalePrice ~", paste(best_model_vars, collapse = " + "))
model_preferred <- lm(as.formula(formula_best), data = ames_clean2)

summary(model_preferred)
```

## Diagnostic Plots - Preferred Model

```{r diagnostics-preferred, fig.height=8}
par(mfrow = c(2, 2))
plot(model_preferred, which = 1, pch = 21, bg = "#b3cccc", col = "#476b6b", cex = 0.8)
plot(model_preferred, which = 2, pch = 21, bg = "#b3cccc", col = "#476b6b", cex = 0.8)
plot(model_preferred, which = 3, pch = 21, bg = "#b3cccc", col = "#476b6b", cex = 0.8)
plot(model_preferred, which = 5, pch = 21, bg = "#b3cccc", col = "#476b6b", cex = 0.8)
par(mfrow = c(1, 1))
```

---

# Model Comparison

```{r final-comparison}
final_comparison <- data.frame(
  Metric = c("Variables", "R-squared", "Adj R-squared", "AIC", "BIC", "RSE"),
  Model_2 = c(
    paste(names(coef(model2))[-1], collapse = ", "),
    round(summary(model2)$r.squared, 4),
    round(summary(model2)$adj.r.squared, 4),
    round(AIC(model2), 2),
    round(BIC(model2), 2),
    round(summary(model2)$sigma, 2)
  ),
  Preferred_Model = c(
    paste(names(coef(model_preferred))[-1], collapse = ", "),
    round(summary(model_preferred)$r.squared, 4),
    round(summary(model_preferred)$adj.r.squared, 4),
    round(AIC(model_preferred), 2),
    round(BIC(model_preferred), 2),
    round(summary(model_preferred)$sigma, 2)
  )
)
print(final_comparison)

if (summary(model_preferred)$adj.r.squared > summary(model2)$adj.r.squared) {
  cat("PREFERRED: All Subsets Model\n")
  cat("Reason: Higher Adjusted R-squared and better model selection criteria\n")
} else {
  cat("PREFERRED: Model 2\n")
  cat("Reason: Similar performance with fewer variables (parsimony principle)\n")
}
```

---

# Conclusion

This study demonstrated a complete regression modeling workflow using the Ames Housing dataset, emphasising both predictive performance and diagnostic validity. Through exploratory data analysis and correlation assessment, key predictors of housing sale prices were identified - overall quality, above-grade living area, garage area, and house age.

An initial regression model using these variables explained approximately **77%** of the variation in sale prices. Comprehensive diagnostic evaluation revealed mild heteroscedasticity and highly influential observations, particularly properties with above-grade living areas exceeding 4,000 sq ft. Removing these improved the adjusted R² to **0.79**.

An automated all subsets regression approach identified a preferred 8-predictor model achieving an adjusted R² of **0.84**, incorporating additional predictors such as basement area, bathroom count, fireplaces, and lot size. The analysis highlights the trade-off between model interpretability and predictive performance, and underscores the importance of diagnostic evaluation and informed model selection in applied regression analysis.

---

# References

- De Cock, D. (2011). Ames, Iowa: Alternative to the Boston housing data as an end of semester regression project. *Journal of Statistics Education, 19*(3). https://doi.org/10.1080/10691898.2011.11889627
- Fox, J., & Weisberg, S. (2019). *An R Companion to Applied Regression* (3rd ed.). Sage Publications.
- James, G., Witten, D., Hastie, T., & Tibshirani, R. (2021). *An Introduction to Statistical Learning with Applications in R* (2nd ed.). Springer.
- Lumley, T. (2020). *leaps: Regression subset selection* (R package version 3.1). https://CRAN.R-project.org/package=leaps
- R Core Team. (2024). *R: A language and environment for statistical computing*. https://www.R-project.org/
- Wei, T., & Simko, V. (2021). *corrplot: Visualization of a correlation matrix* (R package version 0.92). https://CRAN.R-project.org/package=corrplot
- Wickham, H., & Grolemund, G. (2017). *R for data science*. O'Reilly Media.

---
